---
title: "Diffusion in a nutshell"
author: "Sergio Picart-Armada"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
    %\VignetteIndexEntry{Quick start}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE, error = FALSE, 
    fig.width = 7, fig.height = 6)
```

## Getting started

`diffusion` is an R package providing several scores 
for diffusion in networks. 
While its original purpose lies on biological networks, 
its usage is not limited to that scope. 
In general terms, `diffusion` builds several propagation algorithms 
on the \code{igraph} package classes and methods. 

To get started, we will load a toy graph included in the package. 

```{r}
library(diffusion)
data("graph_toy")
```

Let's take a look in the graph:

```{r}
graph_toy
plot(graph_toy)
```

In the next section, we will be running diffusion algorithms 
on this tiny lattice graph. 

## Specifying the input

The package `diffusion` is flexible and allows 
several inputs at once for a given network. 
The input format is, in its most general form, 
a list of matrices, where each matrix contains 
measured nodes in rows and specific scores in columns. 
**Differents sets of scores may have different backgrounds**, 
meaning that we can specifically tag sets of nodes as **unlabelled**. 
If we dispose of a unique list of nodes for label propagation, 
we should provide a list with a unique column vector 
that contains `1`'s in the labels in the list and `0`'s otherwise.

In this example data, the graph contains one input already. 
```{r}
input_vec <- graph_toy$input_vec

head(input_vec, 15)
```

Let's check how many nodes have values

```{r}
length(input_vec)
```

We see that all the nodes have a measure in each of the four score sets. 
In practice, these score sets could be disease genes, pathways, et cetera.

## The diffusion algorithm

Each one of these columns in the input can be *smoothed* using the network 
and new value will be derived - unlabelled nodes are also scored. 
This is the main purpose of diffusion: to derive new scores that 
intend to keep the same trends as the scores in the input, 
but taking into account the network structure.

Let's start with the simplest case of diffusion: 
only a vector of values is to be smoothed. 
Note that these 
**values must be named and must be a subset or all of the graph nodes**.

```{r}
output_vec <- diffusion::diffuse(
    graph = graph_toy, 
    method = "raw", 
    scores = input_vec)

head(output_vec, 15)
```

## Diffusion scores visualisation

The best way to visualise the scores is overlaying 
them in the original lattice. 
`diffusion` also comes with basic mapping functions 
for graphical purposes. 
Let's see an example:

```{r}
igraph::plot.igraph(
    graph_toy, 
    vertex.color = diffusion::scores2colours(output_vec),
    vertex.shape = diffusion::scores2shapes(input_vec),
    main = "Diffusion scores in our lattice"
)
```

Here, we have mapped the scores to colours using `scores2colours` 
and we have highlighted the nodes that were in the 
original input using `scores2shapes` on the original scores.
Square nodes were labelled as relevant in the input, 
and the diffusion algorithm smoothed these labels over the network - 
as in the guilt-by-association principle.

## Several inputs, several smoothing scores

The input to `diffuse` can be more than a vector with scores. 
It can be provided with a set of score vectors, stored in a matrix 
by columns, where rownames should contain 
the nodes that are being scored. 
As different score sets might have different labelled/unlabelled nodes,
`diffuse` also accepts a list of score matrices that may have 
a different amount of rows.

In this section, we will diffuse using a matrix of scores that 
contains four sets of scores, with four different names. 
These example names refer to what the input contains:

* Single: a single node is labelled as positive
* Row: a row of nodes in the lattice graph are positives
* Small_sample: a randomly generated small sample of the lattice nodes 
are positives
* Large_sample: a randomly generated sample with half of the lattice nodes 
are positives

```{r}
input_mat <- graph_toy$input_mat

head(input_mat)
```

On the other hand, there are a variety of methods 
to compute the diffusion scores. 
At the moment, the following: `raw`, `ml` and `gm` for 
classical propagation; `z` and `mc` for scores normalised 
through a statistical model, and similarly `ber_s` and `ber_p`, 
as described in [@mosca]. 
The scoring methods `mc` and `ber_p` require permutations -thus being computationally intense- 
whereas the rest are deterministic. 

For instance, let's smooth through `mc` the input matrix:

```{r}
output_mc <- diffusion::diffuse(
    graph = graph_toy, 
    method = "mc", 
    scores = input_mat)

head(output_mc)
```

We can plot the result of the fourth column *Large_sample*:

```{r}
score_col <- 4
igraph::plot.igraph(
    graph_toy, 
    vertex.color = diffusion::scores2colours(output_mc[, score_col]),
    vertex.shape = diffusion::scores2shapes(input_mat[, score_col]),
    main = "Diffusion scores in our lattice"
)
```

Each method has its particularities and, in the end, 
it is all about the question being asked to the data and 
the particularities of the dataset. 

## Benchmarking

Package `diffusion` offers the option to assess the performance
of the diffusion scores given user-defined target scores or labels. 

The validation must be supplied with the same format as 
the input scores, but 
the labels of the nodes might be different. 
For example, we can diffuse labels on all the nodes of a 
graph but evaluate using only a specific subset of nodes 
and target labels. 
A small example: we want to evaluate how good the diffusion scores 
`raw` and `ml` are at recovering the original labels of the 
first 15 nodes 
when diffusing in the example network. 

```{r}
df_perf <- perf(
    graph = graph_toy,
    scores = graph_toy$input_mat,
    validation = graph_toy$input_mat[1:15, ],
    grid_param = expand.grid(method = c("raw", "ml")))
df_perf
```

This indicates that both methods have a very high 
area under the curve in this example: the ordering of 
the diffusion scores is very aligned to the class label.


## References
